---
title: "R_microbiome_analysis_final"
author: "Vienna"
date: "4/25/2020"
output: 
  pdf_document: default
  keep_md: true
  html_document: default
self_contained: no
thumbnails: no
keep_md: yes
---

```{r setup, include = FALSE}
library(RSQLite)
library(parallel)
library(DECIPHER)
library(gridExtra)
library(ggnewscale) ##new_scale()
library(ggpubr) # ggarrange()
library(ggthemes)
library(grid) # rasterGrob()
library(vegan)
library(phyloseq)
library(png)
library(qgraph) # graph layout
library(ggsignif)
library(RColorBrewer)
library(tidyverse) # includes: ggplot2, dplyr, tidyr, readr, purr, tibble, stringr, forcats
library(knitr)

knitr::opts_chunk$set(echo = FALSE
                      , comment = NA
                      , warning = FALSE
                      , error = FALSE
                      , message = FALSE
                      , tidy = TRUE)
set.seed(42)
rm(list = ls())

# Plotting options
plot_res = 300 # DPI of PNG graphs
col_cembio <- c(rgb(181, 6, 155, 255, max = 255), rgb(64, 118, 253, 255, max = 255))
col_c12 <- brewer.pal(12, "Set3")
theme_set(theme_bw())

# Analysis switches
do_tech_analysis <- TRUE
```


## Summary
The experiment aimed to study the response of honey bee microbiome to Chlorothalonil across host generations and to detect potential microbiome mediated effects on host phenotypes. 
## Methods
We reared adult bees under controlled lab conditions and inoculated them with a natural bee microbiome (start_pool). Bees were orally stressed with Chloropthalonil or no stressor (control). Three cages per treatment were used. Both, the stress-exposed and control microbiomes were transferred to the next cycle (cage to cage transfer) which was handled as before. The third cycle aims to study effects of the pre-exposed microbiome on phenotypes of naive bee hosts in comparison to control microbiomes. For that we transferred the microbiomes and let them be established in the bee hosts without any stress factor contact. We finally applied high amounts of the chemical to all bees with a control microbiome or a pre-exposed microbiome. Bee samples for 16S sequencing has been snap-frozen after cycle 1, cycle 2, cycle 3 BEFORE high stress and cycle 3 AFTER high stress. In addition, the macerated gut pools for transferring microbiome has been saved (start_microbiome as well as pool from each cage to cage transfer). We sequenced the V3-V4 region of the 16S region. Two DNA mock samples from ZymoResearch have been sequenced and named as "positive_control".

### Data and metadata

Overview over all experimental variables:

**Experimental variables find in metadata file**:  

*sample_type* <- microbiome_transfer or single bee
*treatment* <- treatment used (control or which toxin) and additional information if microbiome transfer (e.g. Control_transfer), single bee (e.g. Control)
*treatment2* <- only treatment, not indicating sample type
*cage* <- cage number (three cages per treatment have been used)
*treatment_cage* <- combined information of treatment and cage number (e.g. Control_2)
*date* <- date of sampling during experiment
*cycle* <- experimental cycle (cycle 1, cycle 2, cycle 3 before stress, cycle 3 after stress)
*treatment_cycle* <- experimental cycle in combination with treatment information
*treatment_cycle2*<- experimental cycle in combination with treatment  plus sample type information


```{r load}

#read in otu table
otu_table=read.csv("data_files/Svtab1.csv",sep=",",row.names=1)
otu_table=as.matrix(otu_table)

#read in taxonomy
taxonomy=read.csv("data_files/taxonomy2.csv",sep=",",row.names=1)
taxonomy <- cbind(taxonomy, ASV = paste0("ASV", sprintf("%04d", 1:nrow(taxonomy))))
taxonomy=as.matrix(taxonomy)

metatable <- read.delim("data_files/metadata_reorder3.txt") 
#View(metatable)
row.names(metatable) <- metatable[[1]]
metatable<- metatable[,(-1)]
META <- sample_data(metatable)

#import as phyloseq objects
OTU=otu_table(otu_table,taxa_are_rows=TRUE)
TAX=tax_table(taxonomy)

sample_names(META) == sample_names(OTU)

#create phyloseq object
ps1<- phyloseq(OTU, TAX, META)

# change taxonomy header 
colnames(tax_table(ps1)) <- c(D0 = "Kingdom", D1 = "Phylum", D2 = "Class", 
                              D3 = "Order", D4 = "Family", D5 = "Genus", D6 = "Species", ASV = "ASV")
#check
rank_names(ps1)

#The total number of ASVs in the whole dataset is 
length(taxa_names(ps1))

#get rid of non-bacterial reads and singletons
ps1 = subset_taxa(ps1, Kingdom =="Bacteria")
ps1 <- prune_taxa(taxa_sums(ps1) > 1, ps1)
ps1<-subset_taxa(ps1, (Order!="Chloroplast"))
ps1<-subset_taxa(ps1, (Family!="Mitochondria"))
length(taxa_names(ps1))
#reduces total taxa numbers from 1717 to 1167 (minus 550)

# mean, max and min of sample read counts
smin <- min(sample_sums(ps1))
smean <- mean(sample_sums(ps1))
smax <- max(sample_sums(ps1))
# printing the results
cat("The minimum sample read count is:",smin)
cat("The average sample read count is:",smean)
cat("The maximum sample read count is:",smax)

# min 14240 mean 27937max 49370

#build a custom color palette for phyloseq object
getPalette = colorRampPalette(brewer.pal(12, "Set3"))
speciesList = unique(tax_table(ps1)[,"Genus"])
speciesPalette = getPalette(length(speciesList))
names(speciesPalette) = speciesList
```

### Plot rarefaction curves
```{r rarefaction_curve, fig.heigt=12, fig.width=9}
rare<- ps1
rare2 = subset_samples(rare, treatment2 != "Mock")
rare2 = subset_samples(rare2, treatment2 != "hive_bee")

smin <- min(sample_sums(rare2))
cat("The minimum sample read count is:",smin)

# rarefy without replacement
set.seed(42)
ps.rarefied = rarefy_even_depth(rare2, rngseed=1, sample.size=14240, replace=F)
ps.rarefied
library(ranacapa)
cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7","maroon","khaki4")
ggrare(ps.rarefied, step = 100, se = FALSE, color="treatment2")+ scale_colour_manual(values=cbPalette)+geom_line(size=1)+ theme_bw()+theme(axis.line = element_line(colour = "black"))+ theme(legend.title = element_blank(),
                legend.background = element_blank(),
                legend.key = element_blank()) + theme(text = element_text(size = 18))+theme(strip.text = element_text(face="bold", size=17))+ theme(axis.title.x=element_blank())+theme(plot.title = element_text(hjust = 0.5, size=18, face="italic"))+theme(legend.text=element_text(size=16))+ylim(20,80)
#ggsave("microbiome/figures/supp_rarefaction_sub.png", height = 5, width = 12)

ggrare(rare2, step = 100, se = FALSE, color="treatment2")+ scale_colour_manual(values=cbPalette)+geom_line(size=1)+ theme_bw()+theme(axis.line = element_line(colour = "black"))+ theme(legend.title = element_blank(),
                legend.background = element_blank(),
                legend.key = element_blank()) + theme(text = element_text(size = 18))+theme(strip.text = element_text(face="bold", size=17))+ theme(axis.title.x=element_blank())+theme(plot.title = element_text(hjust = 0.5, size=18, face="italic"))+theme(legend.text=element_text(size=16))+ylim(20,80)
#ggsave("microbiome/figures/supp_rarefaction_no_sub.png", height = 5, width = 12)
```


*** Supplementary figure: Rarefaction curves were used as a qualitative method to estimate the species richness as a function of sequencing depth for all samples. Rarefaction curves reached their asymptotes for all samples, suggesting that saturation in sequencing was achieved.

### Positive controls / mock community

The experiment contained two positive control samples with DNA from the ZymoBIOMICS™ Microbial
Community Standard (Zymo Research).


#### Plot taxa in Mock control samples

```{r mock}
ps_mock <- ps1
ps_mock <- subset_samples(ps_mock, treatment == "Mock")
ps_mock <- prune_taxa(taxa_sums(ps_mock) > 0, ps_mock)
#transform to proportional data
ps_mock2 <- transform_sample_counts(ps_mock, function(OTU) {OTU / sum(OTU)})
ps_mock2 <- tax_glom(ps_mock2, taxrank = 'Genus')
low1pc_reads <- max(taxa_sums(ps_mock2)) * 1 / 100
low1pc_indices <- which(taxa_sums(ps_mock2) < low1pc_reads)
length(low1pc_indices)
taxa_names(ps_mock2)[low1pc_indices[1]] <- "other"
ps_merge <- merge_taxa(ps_mock2, low1pc_indices, "other")
tax_table(ps_merge)["other", ] <- c("other")

getPalette = colorRampPalette(brewer.pal(10, "Set3"))
speciesList = unique(tax_table(ps_merge)[,"Genus"])
speciesPalette = getPalette(length(speciesList))
names(speciesPalette) = speciesList

ps_df <- psmelt(ps_merge)

ps_df_sum <- ps_df %>%
  group_by(Sample,Genus) %>%
  summarise(Abundance = mean(Abundance)) %>%
  group_by(Sample) %>%
  mutate(Relative_abundance = Abundance / sum(Abundance)) %>%
  ungroup() %>%
  mutate(Genus = fct_reorder(Genus, Abundance, .fun = sum, .desc = TRUE)) %>%
  droplevels()
levels(ps_df_sum$Genus)

mock_plot<-ggplot(data = ps_df_sum, aes(x = Sample, y = Relative_abundance, fill = Genus)) +
  geom_bar(stat = "identity")+ scale_y_continuous(expand = c(0,0))+ scale_fill_manual(values= speciesPalette)+ylab("relative abundance")
mock_plot<-mock_plot+ theme(legend.text=element_text(size=9,face="italic"))
mock_plot<-mock_plot+theme(axis.text.x = element_text(size=12, face="bold"))+ theme(axis.title.x=element_blank())+ theme(legend.title=element_blank())+theme(axis.text.y = element_text(size=12, face="bold"))+ theme(axis.title.y=element_text(size=13, face="bold"))
mock_plot2<-mock_plot+theme(plot.title = element_text(face = "bold", color="grey40",size = (17)))+theme(plot.title = element_text(hjust = 0.5))
ps_df_sum
#ggsave("microbiome/figures/mock.png", height = 3, width = 5)
```

All 8 Mock bacterial taxa correctly detected. 
**Figure S1: Comparison of the mock community with the theoretical proportions.**
The two mock samples sequenced in this study **(A)** show no big difference to the expected
theoretical proportions of the reference **(B, bar with label *theoretical*)**. The three other bacterial taxa belong to honey bee core microbiome (Bartonella, Gilliamella and Snodgrassella). The relative abundance of these three taxa across the two Mock samples account for 0.23% of all reads, representing neglectable cross-contamination during sequencing.



#### comparing difference between methods
We sequenced two types of samples: whole bee abdomen and the mix of three macerated guts for cage to cage transfers after each cycle
Therefore, we need to test if these samples are different due to the differences in the methods prior extracting before deciding if we include all or not.
Use PERMANOVA on bray-curtis dissimilarities using proportional transformed abundance data
```{r}
set.seed(42)
methods = ps1
methods <- transform_sample_counts(methods, function(OTU) {OTU / sum(OTU)})

Control = subset_samples(methods, treatment2 == "Control")
Control <- prune_taxa(taxa_sums(Control) > 0, Control)
C_metadata <- as(sample_data(Control), "data.frame")
adonis2(phyloseq::distance(Control, method="bray") ~ sample_type,data = C_metadata,perm=999)

Chloro = subset_samples(methods, treatment2 == "Chlorothalonil")
Chloro <- prune_taxa(taxa_sums(Chloro) > 0, Chloro)
C_metadata <- as(sample_data(Chloro), "data.frame")
adonis2(phyloseq::distance(Chloro, method="bray") ~ sample_type,data = C_metadata, perm=999)
```
No significant difference between gut pools or whole bees and therefore, we continue with analysing them combined


##  Alpha diversity 

```{r alpha_Shannon, fig.heigt=13, fig.width=8}
alpha<- ps1
#pick cycle 1, 2 and 3 before stress
alpha1 = subset_samples(alpha, cycle=="cycle_one" | cycle=="cycle_two" | cycle == "cycle_three_before_stress")
#rarefy to even numbers
set.seed(1)  
alpha1_ra <- rarefy_even_depth(alpha1,sample.size=14240, replace=FALSE, rngseed = 1) 
alpha1_ra<-prune_taxa(taxa_sums(alpha1_ra) > 0, alpha1_ra)

sample_data(alpha1_ra)$cycle<-factor(sample_data(alpha1_ra)$cycle,levels=c("cycle_one","cycle_two","cycle_three_before_stress"), labels=c("Cycle one","Cycle two","Cycle three"))
levels(sample_data(alpha1_ra)$cycle)

sigFunc = function(x){
  if(x < 0.001){"***"} 
  else if(x < 0.01){"**"}
  else if(x < 0.055){"*"}
  else{NA}}

alpha_meas = c("Shannon")
p2 <- plot_richness(alpha1_ra,"treatment2",color="treatment2", measures=alpha_meas)
p2$layers <- p2$layers[-1]
p2<-p2 + geom_boxplot(data=p2$data, aes(x=treatment2, y=value),show.legend=FALSE, alpha=0.1)+geom_point(position = position_jitter(w = 0.2, h = 0.1),show.legend=FALSE) +
  geom_signif(comparisons=list(c("Control", "Chlorothalonil")),tip_length = 0.02,vjust=0.5, color=("grey20"),map_signif_level=sigFunc, y_position = c(3.7,3.7,3.9), margin_top = 0.1, textsize=6, test = wilcox.test)
p2 <- p2 + guides(fill=guide_legend(title=element_blank()))+ylim(2.3,4)
p2 <- p2 + theme(plot.title = element_text(hjust = 0.5))+scale_color_manual(values=c("#55596a", "cyan4"))
p2 <- p2+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+ theme(panel.background = element_blank())
p2 <- p2+theme(axis.line = element_line(colour = "black"))
p2<- p2 + theme(legend.title = element_blank(),legend.background = element_blank(),legend.key = element_blank()) +ylab("Shannon alpha diversity")
p2 <- p2+ theme(panel.border = element_blank())
p2<-p2 + theme(text = element_text(size = 19))+theme(strip.text = element_text(face="bold", size=19, color="gray40"))+ theme(axis.title.x=element_blank())+theme(axis.text.x = element_blank())
Shannon2<-p2+ facet_grid( ~cycle,scales="free_x")+theme(axis.ticks.x=element_blank())+theme(legend.position="bottom")+theme(strip.background =element_rect(fill="white", color="white"))
Shannon2<-Shannon2+theme(panel.spacing = unit(1.1, "cm"))
Shannon2
#ggsave("microbiome/figures/alpha_Shannon2.png", height = 3.5, width = 6)
```
***The Shannon index  captures  information  about  both  the  species  richness  (number  of species) and the relative abundances of the species. Increased Shannon index means increased species diversity.

### make statistical comparisons on Shannon alpha diversity index - comparing treatments against respective controls in each cycle

```{r warning=FALSE}
alpha<- ps1
alpha1 = subset_samples(alpha, cycle=="cycle_one" | cycle=="cycle_two" | cycle == "cycle_three_before_stress")
#rarefy to even numbers
set.seed(1)  
alpha1_ra <- rarefy_even_depth(alpha1,sample.size=14240, replace=FALSE, rngseed = 1) 
alpha1_ra<-prune_taxa(taxa_sums(alpha1_ra) > 0, alpha1_ra)

results = estimate_richness(alpha1_ra, measures = 'Shannon')
d = sample_data(alpha1_ra)

# calculate wilcox-test
Control_1 = results[d[,'treatment_cycle'] == 'Control_cycle_1',]
Chlorothalonil_1 = results[d[,'treatment_cycle'] == 'Chlorothalonil_cycle_1',]
wilcox.test(Control_1, Chlorothalonil_1)
#W = 106, p-value = 0.052

#cycle 2
Control_2 = results[d[,'treatment_cycle'] == 'Control_cycle_2',]
Chlorothalonil_2 = results[d[,'treatment_cycle'] == 'Chlorothalonil_cycle_2',]
wilcox.test(Control_2, Chlorothalonil_2)
#W = 63, p-value = 0.6297

#cycle 3
control_cycle_3_before_stress = results[d[,'treatment_cycle'] == 'control_Chlorothalonil_cycle_3_before_stress',]
Chlorothalonil_cycle_3_before_stress = results[d[,'treatment_cycle'] == 'Chlorothalonil_cycle_3_before_stress',]
wilcox.test(control_cycle_3_before_stress, Chlorothalonil_cycle_3_before_stress)
#W = 30, p-value = 0.3865
```
Under chlorothalonil the Shannon alpha diversity was significantly affected only in cycle 1 


```{r alpha_Observed, fig.heigt=13, fig.width=8}
alpha<- ps1
#pick cycle 1, 2 and 3 before stress
alpha1 = subset_samples(alpha, cycle=="cycle_one" | cycle=="cycle_two" | cycle == "cycle_three_before_stress")
#rarefy to even numbers
set.seed(1)  
alpha1_ra <- rarefy_even_depth(alpha1,sample.size=14240, replace=FALSE, rngseed = 1) 
alpha1_ra<-prune_taxa(taxa_sums(alpha1_ra) > 0, alpha1_ra)

sample_data(alpha1_ra)$cycle<-factor(sample_data(alpha1_ra)$cycle,levels=c("cycle_one","cycle_two","cycle_three_before_stress"), labels=c("Cycle one","Cycle two","Cycle three"))
levels(sample_data(alpha1_ra)$cycle)

sigFunc = function(x){
  if(x < 0.001){"***"} 
  else if(x < 0.01){"**"}
  else if(x < 0.05){"*"}
  else{NA}}

alpha_meas2 = c("Observed")
p2 <- plot_richness(alpha1_ra,"treatment2",color="treatment2", measures=alpha_meas2)
p2$layers <- p2$layers[-1]
p2<-p2 + geom_boxplot(data=p2$data, aes(x=treatment2, y=value),show.legend=FALSE, alpha=0.1)+geom_point(position = position_jitter(w = 0.2, h = 0.1),show.legend=FALSE) +
  geom_signif(comparisons=list(c("Control", "Chlorothalonil")),tip_length = 0.02,vjust=0.5, color=("grey20"),map_signif_level=sigFunc, margin_top = 0.1, textsize=6, test = wilcox.test)
p2 <- p2 + guides(fill=guide_legend(title=element_blank()))
p2 <- p2 + theme(plot.title = element_text(hjust = 0.5))+scale_color_manual(values=c("#55596a", "cyan4"))
p2 <- p2+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+ theme(panel.background = element_blank())
p2 <- p2+theme(axis.line = element_line(colour = "black"))
p2<- p2 + theme(legend.title = element_blank(),legend.background = element_blank(),legend.key = element_blank()) +ylab("Observed taxa")
p2 <- p2+ theme(panel.border = element_blank())
p2<-p2 + theme(text = element_text(size = 19))+theme(strip.text = element_text(face="bold", size=19, color="gray40"))+ theme(axis.title.x=element_blank())#+theme(axis.text.x = element_blank())
Observed<-p2+ facet_grid( ~cycle,scales="free_x")+theme(axis.ticks.x=element_blank())+theme(legend.position="bottom")+theme(strip.background =element_rect(fill="white", color="white"))
Observed<-Observed+theme(panel.spacing = unit(1.1, "cm"))
Observed
#ggsave("microbiome/figures/alpha_Observed2.png", height = 3.5, width = 6)
```

*** Supplementary figure Observed alpha diversity
The observed plot shows the difference in detected OTUs. 

### make statistical comparisons on Observed species numbers - comparing treatments against respective controls in each cycle
```{r warning=FALSE}
alpha<- ps1
alpha1 = subset_samples(alpha, cycle=="cycle_one" | cycle=="cycle_two" | cycle == "cycle_three_before_stress")
#rarefy to even numbers
set.seed(1)  
alpha1_ra <- rarefy_even_depth(alpha1,sample.size=14240, replace=FALSE, rngseed = 1) 

results = estimate_richness(alpha1_ra, measures = 'Observed')
d = sample_data(alpha1_ra)

# calculate wilcox-test
Control_1 = results[d[,'treatment_cycle'] == 'Control_cycle_1',]
Chlorothalonil_1 = results[d[,'treatment_cycle'] == 'Chlorothalonil_cycle_1',]
wilcox.test(Control_1, Chlorothalonil_1)
#W = 99, p-value = 0.1227

#cycle 2
Control_2 = results[d[,'treatment_cycle'] == 'Control_cycle_2',]
Chlorothalonil_2 = results[d[,'treatment_cycle'] == 'Chlorothalonil_cycle_2',]
wilcox.test(Control_2, Chlorothalonil_2)
#W = 55, p-value = 0.3388

#cycle 3
control_cycle_3_before_stress = results[d[,'treatment_cycle'] == 'control_Chlorothalonil_cycle_3_before_stress',]
Chlorothalonil_cycle_3_before_stress = results[d[,'treatment_cycle'] == 'Chlorothalonil_cycle_3_before_stress',]
wilcox.test(control_cycle_3_before_stress, Chlorothalonil_cycle_3_before_stress)
#W = 24, p-value = 0.1543
```
Numbers of observed species is not significantly different under chlorothalonil to the control in all three cycles.



# Beta diversity

### ordination plots
```{r NMDS_man, fig.heigt=7.5, fig.width=10}
ord2<- ps1
ord2 = subset_samples(ord2, cycle=="cycle_one" | cycle=="cycle_two" | cycle == "cycle_three_before_stress")

ord2 <- transform_sample_counts(ord2, function(OTU) {OTU / sum(OTU)})
sample_data(ord2)$cycle<-factor(sample_data(ord2)$cycle,levels=c("cycle_one","cycle_two", "cycle_three_before_stress"), labels=c("Cycle one","Cycle two","Cycle three"))
levels(sample_data(ord2)$cycle)

ord2 <- prune_taxa(taxa_sums(ord2) > 0, ord2)
set.seed(42)
dist<-phyloseq::distance(ord2, method="bray") 
ordination = ordinate(ord2, method="NMDS", distance=dist)
cat("stress is:", ordination$stress)
p2 <- plot_ordination(ord2, ordination, color="treatment2") +geom_point(size=2.5)
p2 <- p2 + guides(fill=guide_legend(title=element_blank()))
p2 <- p2 + theme(plot.title = element_text(hjust = 0.5))+scale_color_manual(values=c("#55596a", "cyan4"),guide = guide_legend(override.aes = list(shape = c(rep(15, 2)),size=7,alpha=0.5)))
p2 <- p2+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+ theme(panel.background = element_blank())
p2 <- p2+theme(axis.line = element_line(colour = "black"))
p2<- p2 + theme(legend.title = element_blank(),legend.background = element_blank(),legend.key = element_blank())
p2 <- p2+ theme(panel.border = element_blank())
p2<-p2 + theme(text = element_text(size = 19))+theme(strip.text = element_blank())
p2<-p2+ facet_wrap( ~cycle,scales="free_x")+theme(legend.position="bottom")+theme(strip.background =element_rect(fill="white", color="white"))
NMDS_main<-p2+stat_ellipse(aes(group = treatment2),size=1.1, alpha=0.5)+ theme(panel.spacing.x=unit(5.5, "lines"))+ theme(axis.title.y = element_text(margin = margin(t = 0, r = 0, b = 0, l = 20)))+ scale_x_continuous(limits = c(-1.1, 1), breaks = seq(-1, 1, by = 1))
NMDS_main
#ggsave("microbiome/figures/NMDS_man.png", height = 5, width = 10)

#stress is: 0.1224216
#(<0.2 is good)
```

```{r PCoA, fig.heigt=7.5, fig.width=10}
ord2<- ps1
ord2 = subset_samples(ord2, cycle=="cycle_one" | cycle=="cycle_two" | cycle == "cycle_three_before_stress")

ord2 <- transform_sample_counts(ord2, function(OTU) {OTU / sum(OTU)})
sample_data(ord2)$cycle<-factor(sample_data(ord2)$cycle,levels=c("cycle_one","cycle_two", "cycle_three_before_stress"), labels=c("Cycle one","Cycle two","Cycle three"))
levels(sample_data(ord2)$cycle)

ord2 <- prune_taxa(taxa_sums(ord2) > 0, ord2)
set.seed(42)
dist<-phyloseq::distance(ord2, method="bray") 
ordination2 = ordinate(ord2, method="PCoA", distance=dist)
p3 <- plot_ordination(ord2, ordination2, color="treatment2") +geom_point(size=2.5)
p3 <- p3 + guides(fill=guide_legend(title=element_blank()))
p3 <- p3 + theme(plot.title = element_text(hjust = 0.5))+scale_color_manual(values=c("#55596a", "cyan4"),guide = guide_legend(override.aes = list(shape = c(rep(15, 2)),size=7,alpha=0.5)))
p3 <- p3+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+ theme(panel.background = element_blank())
p2 <- p2+theme(axis.line = element_line(colour = "black"))
p2<- p2 + theme(legend.title = element_blank(),legend.background = element_blank(),legend.key = element_blank())
p2 <- p2+ theme(panel.border = element_blank())
p2<-p2 + theme(text = element_text(size = 16.5))+theme(strip.text = element_blank())
p2<-p2+ facet_wrap( ~cycle,scales="free_x")+theme(legend.position="bottom")+theme(strip.background =element_rect(fill="white", color="white"))
PCoA_main<-p2+stat_ellipse(aes(group = treatment2),size=1.1, alpha=0.5)+ theme(panel.spacing.x=unit(5.5, "lines"))+ theme(axis.title.y = element_text(margin = margin(t = 0, r = 0, b = 0, l = 20)))+ scale_x_continuous(limits = c(-0.6, 0.45), breaks = seq(-0.6, 0.4, by = 0.5))
PCoA_main
#ggsave("microbiome/figures/PCoA.png", height = 5, width = 10)
```


## Beta diversity stats

#### test difference between treatments and respective controls in the three cycles to statistically verify microbial community compositional difference seen in ordination plots 
```{r}
set.seed(42)
compare=ps1
compare1 <- transform_sample_counts(compare, function(OTU) {OTU / sum(OTU)})

cycle1 = subset_samples(compare1, cycle == "cycle_one")
cycle1 <- prune_taxa(taxa_sums(cycle1) > 0, cycle1)
cycle1.1 <- as(sample_data(cycle1), "data.frame")

cycle2 = subset_samples(compare1, cycle == "cycle_two")
cycle2 <- prune_taxa(taxa_sums(cycle2) > 0, cycle2)
cycle2.1 <- as(sample_data(cycle2), "data.frame")

cycle3b = subset_samples(compare1, cycle == "cycle_three_before_stress")
cycle3b <- prune_taxa(taxa_sums(cycle3b) > 0, cycle3b)
cycle3b.1 <- as(sample_data(cycle3b), "data.frame")

#cycle 1

subs <- subset_samples(cycle1, treatment2 %in% c("Control", "Chlorothalonil"))
metadata <- as(sample_data(subs), "data.frame")
adonis2(phyloseq::distance(subs, method="bray") ~ treatment2,data = metadata, perm=999)
#p=0.234
#cycle 2

subs <- subset_samples(cycle2, treatment2 %in% c("Control", "Chlorothalonil"))
metadata <- as(sample_data(subs), "data.frame")
adonis2(phyloseq::distance(subs, method="bray") ~ treatment2,data = metadata, perm=999)
#p=0.5

#cycle 3 before stress

subs <- subset_samples(cycle3b, treatment2 %in% c("Control", "Chlorothalonil"))
metadata <- as(sample_data(subs), "data.frame")
adonis2(phyloseq::distance(subs, method="bray") ~ treatment2,data = metadata, perm=999)
#p=0.133
```
Community composition was not significantly affected under Chlorothalonil in all cycles. 


### Test for cage effects
```{r cage_effects}
set.seed(42)

cycle1_co = subset_samples(cycle1, treatment2 == "Control")
cycle1_co.1 <- as(sample_data(cycle1_co), "data.frame")
d = phyloseq::distance(cycle1_co, "bray")
adonis2(d ~ cage, cycle1_co.1, perm=999)
#p=0.5

cycle1_ch = subset_samples(cycle1, treatment2 == "Chlorothalonil")
cycle1_ch.1 <- as(sample_data(cycle1_ch), "data.frame")
d = phyloseq::distance(cycle1_ch, "bray")
adonis2(d ~ cage, cycle1_ch.1, perm=999)
#p=0.35

#cycle2

cycle2_control = subset_samples(cycle2, treatment2 == "Control")
cycle2_control.1 <- as(sample_data(cycle2_control), "data.frame")
d = phyloseq::distance(cycle2_control, "bray")
adonis2(d ~ cage, cycle2_control.1, perm=999)
#p<0.001

cycle2_ch = subset_samples(cycle2, treatment2 == "Chlorothalonil")
cycle2_ch.1 <- as(sample_data(cycle2_ch), "data.frame")
d = phyloseq::distance(cycle2_ch, "bray")
adonis2(d ~ cage, cycle2_ch.1, perm=999)
#p=0.37

#cycle 3 before stress

cycle3_control = subset_samples(cycle3b, treatment2 == "Control")
cycle3_control.1 <- as(sample_data(cycle3_control), "data.frame")
d = phyloseq::distance(cycle3_control, "bray")
adonis2(d ~ cage, cycle3_control.1, perm=999)
#p=0.32

cycle3_ch = subset_samples(cycle3b, treatment2 == "Chlorothalonil")
cycle3_ch.1 <- as(sample_data(cycle3_ch), "data.frame")
d = phyloseq::distance(cycle3_ch, "bray")
adonis2(d ~ cage, cycle3_ch.1, perm=999)
#p=0.32

#cage differences general (across all cycles)
ord2<- ps1
all = subset_samples(ord2, cycle=="cycle_one" | cycle=="cycle_two" | cycle == "cycle_three_before_stress")
all <- prune_taxa(taxa_sums(all) > 0, all)
all1 <- as(sample_data(all), "data.frame")
d = phyloseq::distance(all, "bray")
adonis2(d ~ treatment_cage, all1, perm=999)
#p=0.06
```

Testing if the three cages within a treatment show significant variations in microbiome composition (cage effects). Only in cycle 2 of the control are significant cage effects visible.


### Time effect
#### Test if microbial composition of treatments differ across the three cycles

```{r}
set.seed(42)
compare = ps1
compare2 <- prune_taxa(taxa_sums(compare) > 0, compare)
compare2 <- transform_sample_counts(compare2, function(OTU) {OTU / sum(OTU)})

subs1 <- subset_samples(compare2, treatment_cycle == "Control_cycle_1"|treatment_cycle=="Control_cycle_2"|treatment_cycle=="control_cycle_3_before_stress")
metadata <- as(sample_data(subs1), "data.frame")
adonis2(phyloseq::distance(subs1, method="bray") ~ cycle,
       data = metadata, perm=999)
#p=0.01

subs4 <- subset_samples(compare2, treatment_cycle %in% c("Chlorothalonil_cycle_1", "Chlorothalonil_cycle_2","Chlorothalonil_cycle_3_before_stress"))
metadata <- as(sample_data(subs4), "data.frame")
adonis2(phyloseq::distance(subs4, method="bray") ~ cycle,
       data = metadata, perm=999)
#p=0.001
```

Treatments differ significantly across cycles

#### compare treatments before and after high stress

```{r}
set.seed(42)
compare2 = ps1
compare2 <- transform_sample_counts(compare2, function(OTU) {OTU / sum(OTU)})

subs1 <- subset_samples(compare2, treatment_cycle %in% c("control_Chlorothalonil_cycle_3_before_stress", "control_Chlorothalonil_cycle_3_after_stress"))
metadata <- as(sample_data(subs1), "data.frame")
adonis2(phyloseq::distance(subs1, method="bray") ~ treatment_cycle,data = metadata, perm=999)
#p=0.22

subs1 <- subset_samples(compare2, treatment_cycle %in% c("Chlorothalonil_cycle_3_before_stress", "Chlorothalonil_cycle_3_after_stress"))
metadata <- as(sample_data(subs1), "data.frame")
adonis2(phyloseq::distance(subs1, method="bray") ~ treatment_cycle,data = metadata, perm=999)
#p=0.22
```

No significant difference for any treatment before and after high stress application


### test spread of variance of groups
```{r betadisp}
set.seed(53)
test<-ps1
test2 <- transform_sample_counts(test, function(OTU) {OTU / sum(OTU)})

cycle1 = subset_samples(test2, cycle == "cycle_one")
cycle1 <- prune_taxa(taxa_sums(cycle1) > 0, cycle1)

cycle2 = subset_samples(test2, cycle == "cycle_two")
cycle2 <- prune_taxa(taxa_sums(cycle2) > 0, cycle2)

cycle3_b = subset_samples(test2, cycle == "cycle_three_before_stress")
cycle3_b <- prune_taxa(taxa_sums(cycle3_b) > 0, cycle3_b)

#testing treatment
set.seed(53)
d_cycle1 = phyloseq::distance(cycle1, "bray")
df_cycle1 = as(sample_data(cycle1), "data.frame")
groups <- df_cycle1[["treatment2"]]
beta <- betadisper(d_cycle1, df_cycle1$treatment2)
permutest(beta,pairwise = TRUE, permutations = 999)
anova(betadisper(d_cycle1, groups))
mod.HSD <- TukeyHSD(beta,conf.level = 0.95)
mod.HSD
plot(betadisper(d_cycle1, groups),main="MultiVariate Permutation Cycle one",lwd=2, label = TRUE,label.cex = 0.9,col=c("#55596a","cyan4"))
boxplot(betadisper(d_cycle1, groups),main="Cycle one",border=c("#55596a","cyan4"),xlab="")

set.seed(53)
d_cycle2 = phyloseq::distance(cycle2, "bray")
df_cycle2 = as(sample_data(cycle2), "data.frame")
groups <- df_cycle2[["treatment2"]]
beta <- betadisper(d_cycle2, df_cycle2$treatment2)
permutest(beta,pairwise = TRUE, permutations = 999)
anova(betadisper(d_cycle2, groups))
mod.HSD <- TukeyHSD(beta)
mod.HSD
plot(betadisper(d_cycle2, groups),main="MultiVariate Permutation Cycle two",lwd=2, label = TRUE,label.cex = 0.9,col=c("#55596a","cyan4"))
boxplot(betadisper(d_cycle2, groups),main="Cycle two",border=c("#55596a","cyan4"),xlab="")

set.seed(53)
d_cycle3_b = phyloseq::distance(cycle3_b, "bray")
df_cycle3_b = as(sample_data(cycle3_b), "data.frame")
groups <- df_cycle3_b[["treatment2"]]
beta <- betadisper(d_cycle3_b, df_cycle3_b$treatment2)
permutest(beta,pairwise = TRUE, permutations = 999)
anova(betadisper(d_cycle3_b, groups))
mod.HSD <- TukeyHSD(beta)
mod.HSD
plot(betadisper(d_cycle3_b, groups),main="MultiVariate Permutation Cycle three",lwd=2, label = TRUE,label.cex = 0.9,col=c("#55596a","cyan4"))
boxplot(betadisper(d_cycle3_b, groups),main="Cycle three",border=c("#55596a","cyan4"),xlab="")
```

# Taxonomy plots and stats

```{r set_up_for_taxa_plots}
tax <-ps1
tax = subset_samples(tax, treatment != "positive_control")
ps_bdiv2 <- transform_sample_counts(tax, function(OTU) {OTU / sum(OTU)})
tax_all2<- ps_bdiv2
ps_pd2 <- tax_glom(tax_all2, taxrank = 'Genus')

low1pc_reads <- max(taxa_sums(ps_pd2)) * 1 / 100
low1pc_indices <- which(taxa_sums(ps_pd2) < low1pc_reads)
length(low1pc_indices)
taxa_names(ps_pd2)[low1pc_indices[1]] <- "other"
taxa3 <- merge_taxa(ps_pd2, low1pc_indices, "other")
tax_table(taxa3)["other", ] <- c("other")

#build a custom color palette for phyloseq object
getPalette = colorRampPalette(brewer.pal(8, "Set3"))
speciesList = unique(tax_table(taxa3)[,"Genus"])
speciesPalette = getPalette(length(speciesList))
names(speciesPalette) = speciesList
```


```{r taxa_cycle1}

ps2 <-taxa3
cycle1 = subset_samples(ps2, cycle=="cycle_one")
cycle1 <- prune_taxa(taxa_sums(cycle1) > 0, cycle1)

ps_df_c1 <- psmelt(cycle1)
ps_df_sum_c1 <- ps_df_c1 %>%
  group_by(treatment2,Genus,cycle,Sample) %>%
  summarise(Abundance = mean(Abundance)) %>%
  group_by(cycle,treatment2,Sample) %>%
  mutate(relative_abundance = Abundance / sum(Abundance)) %>%
  ungroup() %>%
  mutate(Genus = fct_reorder(Genus, Abundance, .fun = sum, .desc = TRUE))
levels(ps_df_sum_c1$Genus)

change <- c("Cycle one")
ps_df_sum_c1 <- arrange(transform(ps_df_sum_c1, cycle=factor(cycle,labels=change)),cycle)

ps_df_sum_c1$Sample <- factor(ps_df_sum_c1$Sample, levels = unique(ps_df_sum_c1$Sample[order(ps_df_sum_c1$treatment2)]))

p<-ggplot(data = ps_df_sum_c1, aes(x = Sample, y = relative_abundance, fill = Genus)) +
  geom_bar(stat = "identity")+ scale_fill_manual(values= speciesPalette)+ scale_y_continuous(expand = c(0,0))+ylab("relative abundance")
p<-p+ theme(legend.text=element_text(size=18, face = "italic"))+theme(legend.key = element_rect(color = NA, fill = NA),
                                                     legend.key.size = unit(0.9, "cm"))+theme(legend.title = element_blank())+theme(axis.ticks.x = element_blank())
p<-p + facet_grid(cycle~treatment2, scales="free_x",space="free_x")
p<-p + theme(strip.text = element_text(size=18,face="bold",color="grey35"))
cycle1_plot<-p+theme(axis.title.y = element_text(size=17, face="bold"))+theme(axis.text.y = element_text(size=15, face="bold"))+theme(axis.text.x = element_text(size=11, face="bold", angle = 90))+theme(axis.title.x = element_blank())+theme(strip.background =element_rect(fill="white",color="white"))+theme(axis.text.x = element_blank())+theme(legend.position="bottom")
#ggsave("microbiome/figures/cycle1_plot.png", height = 6, width = 10)
```

```{r taxa_cycle2}

ps2 <-taxa3
cycle2 = subset_samples(ps2, cycle=="cycle_two")
cycle2 <- prune_taxa(taxa_sums(cycle2) > 0, cycle2)

ps_df_c2 <- psmelt(cycle2)
ps_df_sum_c2 <- ps_df_c2 %>%
  group_by(treatment2,Genus,cycle,Sample) %>%
  summarise(Abundance = mean(Abundance)) %>%
  group_by(cycle,treatment2,Sample) %>%
  mutate(relative_abundance = Abundance / sum(Abundance)) %>%
  ungroup() %>%
  mutate(Genus = fct_reorder(Genus, Abundance, .fun = sum, .desc = TRUE))
levels(ps_df_sum_c2$Genus)

change <- c("Cycle two")
ps_df_sum_c2 <- arrange(transform(ps_df_sum_c2, cycle=factor(cycle,labels=change)),cycle)

ps_df_sum_c2$Sample <- factor(ps_df_sum_c2$Sample, levels = unique(ps_df_sum_c2$Sample[order(ps_df_sum_c2$treatment2)]))

p<-ggplot(data = ps_df_sum_c2, aes(x = Sample, y = relative_abundance, fill = Genus)) +
  geom_bar(stat = "identity")+ scale_fill_manual(values= speciesPalette)+ scale_y_continuous(expand = c(0,0))+ylab("relative abundance")
p<-p+ theme(legend.text=element_text(size=18, face = "italic"))+theme(legend.key = element_rect(color = NA, fill = NA),
                                                     legend.key.size = unit(0.9, "cm"))+theme(legend.title = element_blank())
#p<-p + facet_wrap(cycle~treatment3, scales="free_x",nrow = 1)
p<-p + facet_grid(cycle~treatment2, scales="free_x",space="free_x")
p<-p + theme(strip.text = element_text(size=18,face="bold",color="grey35"))
cycle2_plot<-p+theme(axis.title.y = element_text(size=18, face="bold"))+theme(axis.text.y = element_text(size=15, face="bold"))+theme(axis.text.x = element_text(size=11, face="bold", angle = 90))+theme(axis.title.x = element_blank())+theme(axis.text.x = element_text(vjust = 0.5,hjust=1))+theme(strip.background =element_rect(fill="white",color="white"))+theme(legend.position="bottom")+ theme(axis.text.x = element_text(hjust = -0.2))+theme(axis.text.x = element_blank())+theme(axis.ticks.x = element_blank())
#ggsave("microbiome/figures/cycle2_plot.png", height = 6, width = 10)
```


```{r taxa_cycle3}

ps2 <-taxa3
cycle3 = subset_samples(ps2, cycle=="cycle_three_before_stress")
cycle3 <- prune_taxa(taxa_sums(cycle3) > 0, cycle3)

ps_df_c3 <- psmelt(cycle3)
ps_df_sum_c3 <- ps_df_c3 %>%
  group_by(treatment2,Genus,cycle,Sample) %>%
  summarise(Abundance = mean(Abundance)) %>%
  group_by(cycle,treatment2,Sample) %>%
  mutate(relative_abundance = Abundance / sum(Abundance)) %>%
  ungroup() %>%
  mutate(Genus = fct_reorder(Genus, Abundance, .fun = sum, .desc = TRUE))
levels(ps_df_sum_c3$Genus)

change <- c("Cycle three")
ps_df_sum_c3 <- arrange(transform(ps_df_sum_c3, cycle=factor(cycle,labels=change)),cycle)

ps_df_sum_c3$Sample <- factor(ps_df_sum_c3$Sample, levels = unique(ps_df_sum_c3$Sample[order(ps_df_sum_c3$treatment2)]))

p<-ggplot(data = ps_df_sum_c3, aes(x = Sample, y = relative_abundance, fill = Genus)) +
  geom_bar(stat = "identity")+ scale_fill_manual(values= speciesPalette)+ scale_y_continuous(expand = c(0,0))+ylab("relative abundance")
p<-p+ theme(legend.text=element_text(size=18, face = "italic"))+theme(legend.key = element_rect(color = NA, fill = NA),
                                                     legend.key.size = unit(0.7, "cm"))+theme(legend.title = element_blank())
#p<-p + facet_wrap(cycle~treatment3, scales="free_x",nrow = 1)
p<-p + facet_grid(cycle~treatment2, scales="free_x",space="free_x")
p<-p + theme(strip.text = element_text(size=18,face="bold",color="grey35"))
cycle3_plot<-p+theme(axis.title.y = element_text(size=17, face="bold"))+theme(axis.text.y = element_text(size=15, face="bold"))+theme(axis.text.x = element_text(size=11, face="bold", angle = 90))+theme(axis.title.x = element_blank())+theme(axis.text.x = element_text(vjust = 0.5,hjust=1))+theme(strip.background =element_rect(fill="white",color="white"))+theme(legend.position="bottom")+ theme(axis.text.x = element_text(hjust = -0.2))+theme(axis.text.x = element_blank())+theme(axis.ticks.x = element_blank())
#ggsave("microbiome/figures/cycle3_plot.png", height = 6, width = 10)
```


# microbiome in bees cycle 3 after high stress application 
```{r taxa_cycle3_after_stress}

ps2 <-taxa3
cycle3.2 = subset_samples(ps2, cycle=="cycle_three_after_stress")
cycle3.2 <- prune_taxa(taxa_sums(cycle3.2) > 0, cycle3.2)

ps_df_c3 <- psmelt(cycle3.2)
ps_df_sum_c3 <- ps_df_c3 %>%
  group_by(treatment2,Genus,cycle,Sample) %>%
  summarise(Abundance = mean(Abundance)) %>%
  group_by(cycle,treatment2,Sample) %>%
  mutate(relative_abundance = Abundance / sum(Abundance)) %>%
  ungroup() %>%
  mutate(Genus = fct_reorder(Genus, Abundance, .fun = sum, .desc = TRUE))
levels(ps_df_sum_c3$Genus)

change <- c("Control","Chlorothalonil")
ps_df_sum_c3 <- arrange(transform(ps_df_sum_c3,treatment2=factor(treatment2,labels=change)),treatment2)

p<-ggplot(data = ps_df_sum_c3, aes(x = Sample, y = relative_abundance, fill = Genus)) +
  geom_bar(stat = "identity")+ scale_fill_manual(values= speciesPalette)+ scale_y_continuous(expand = c(0,0))+ylab("relative abundance")
p<-p+ theme(legend.text=element_text(size=18, face = "italic"))+theme(legend.key = element_rect(color = NA, fill = NA),
                                                     legend.key.size = unit(0.7, "cm"))+theme(legend.title = element_blank())
p<-p + facet_grid(~treatment2, scales="free_x",space="free_x")
p<-p + theme(strip.text = element_text(size=18,face="bold",color="grey35"))
cycle3_plot_after<-p+theme(axis.title.y = element_text(size=17, face="bold"))+theme(axis.text.y = element_text(size=15, face="bold"))+theme(axis.text.x = element_text(size=11, face="bold", angle = 90))+theme(axis.title.x = element_blank())+theme(axis.text.x = element_text(vjust = 0.5,hjust=1))+theme(strip.background =element_rect(fill="white",color="white"))+theme(legend.position="bottom")+ theme(axis.text.x = element_text(hjust = -0.2))+theme(axis.text.x = element_blank())+theme(axis.ticks.x = element_blank())
#ggsave("microbiome/figures/cycle3_after_plot.png", height = 6, width = 9)
```
#nurse bees from same hive as natural control
```{r taxa_nurse}

ps2 <-taxa3
nurse = subset_samples(ps2, treatment=="hive_bee")
nurse <- prune_taxa(taxa_sums(nurse) > 0, nurse)

ps_df_c3 <- psmelt(nurse)
ps_df_sum_c3 <- ps_df_c3 %>%
  group_by(Genus,Sample) %>%
  summarise(Abundance = mean(Abundance)) %>%
  group_by(Sample) %>%
  mutate(relative_abundance = Abundance / sum(Abundance)) %>%
  ungroup() %>%
  mutate(Genus = fct_reorder(Genus, Abundance, .fun = sum, .desc = TRUE))
levels(ps_df_sum_c3$Genus)

p<-ggplot(data = ps_df_sum_c3, aes(x = Sample, y = relative_abundance, fill = Genus)) +
  geom_bar(stat = "identity")+ scale_fill_manual(values= speciesPalette)+ scale_y_continuous(expand = c(0,0))+ylab("relative abundance")
p<-p+ theme(legend.text=element_text(size=13, face = "italic"))+theme(legend.key = element_rect(color = NA, fill = NA),
                                                     legend.key.size = unit(0.55, "cm"))+theme(legend.title = element_blank())
nurse<-p+theme(axis.title.y = element_text(size=15, face="bold"))+theme(axis.text.y = element_text(size=13, face="bold"))+theme(axis.title.x = element_blank())+theme(axis.text.x = element_text(vjust = 0.5,hjust=1))+theme(axis.text.x = element_blank())+theme(axis.ticks.x = element_blank())
#ggsave("microbiome/figures/nurse.png", height = 3.5, width = 6)
```

## Core taxa abundance

#extract core taxa (rarefy to even depth) and plot abundance across treatments and cycles

```{r total abundance}
ps2 <-ps1
ps2 <- rarefy_even_depth(ps2,sample.size=14240, replace=FALSE, rngseed = 1) 
ps2 = subset_samples(ps2, cycle == "cycle_one"| cycle == "cycle_two"| cycle == "cycle_three_before_stress")
ps3 <- prune_taxa(taxa_sums(ps2) > 0, ps2)
ps3 <- tax_glom(ps3, taxrank = 'Genus')
psOrd3 = subset_taxa(ps3, Genus=="Lactobacillus" | Genus=="Bartonella" | Genus=="Gilliamella" | Genus=="Snodgrassella" | Genus=="Frischella" | Genus=="Bifidobacterium" | Genus=="Commensalibacter")

#Melt and plot
melt<-psmelt(psOrd3)
levels2=c("cycle_one","cycle_two","cycle_three_before_stress")
change <- c("Cycle one","Cycle two","Cycle three")
melt2 <- arrange(transform(melt, cycle=factor(cycle,levels=levels2,labels=change)),cycle)
neworder2 <- c("Bartonella","Lactobacillus","Gilliamella","Snodgrassella","Bifidobacterium","Commensalibacter","Frischella")
melt4 <- arrange(transform(melt2, Genus=factor(Genus,levels=neworder2)),Genus)

a_mean <- melt4 %>% 
  group_by(treatment2,Genus,cycle) %>% 
  summarize(mean_val = mean(Abundance))
a_mean2<-subset(a_mean, treatment2 =="Control")
print(a_mean2)

sigFunc = function(x){
  if(x < 0.001){"***"} 
  else if(x < 0.01){"**"}
  else if(x < 0.05){"*"}
  else{NA}}

p<-ggplot(data = melt4, aes(x = treatment2, y = Abundance)) +
  geom_boxplot(aes(fill=Genus),alpha=0.5,lwd=0.7, position = position_dodge(width = 0.3), width=0.45,outlier.shape = NA) + geom_jitter(aes(color = Genus), height = 0, width = .1)+geom_hline(data= a_mean2, aes(yintercept=mean_val), width=5,size=1,linetype="dashed")+
  labs(x = "", y = "Abundance\n")+
  facet_grid(Genus~cycle, scales = "free")+theme_bw()+
  geom_signif(comparisons=list(c("Control", "Chlorothalonil")),tip_length = 0.03,vjust=1.3, color="grey20", size=1.1,map_signif_level=sigFunc, textsize=9, test = wilcox.test,step_increase = 0.2)
p<-p+ theme(legend.position="right")+ylab("Total abundance")
p<-p+ theme(legend.text=element_text(size=29, face = "italic"))+theme(legend.key = element_rect(color = NA, fill = NA),legend.key.size = unit(0.9, "cm"))+theme(legend.title = element_blank())
abu<-p +theme(strip.text.x = element_text(size=25,face="bold",color = "grey 35"))+theme(strip.background =element_rect(fill="white", color="white"))
cycle<-abu+theme(axis.title.y = element_text(size=25, face="bold"))+theme(axis.text.y = element_text(size=20, face="bold"))+theme(axis.text.x = element_text(size=25, angle=90,face="bold"))+theme(axis.title.x = element_blank())+ theme(strip.text.y = element_blank())
#ggsave("microbiome/figures/core_total_abundance.png", height = 20, width = 14)
``` 

### stats on abundances

### compare if core taxa abundances are significantly different between controls and treatments in cycle 1, 2 and cycle 3, use Wilcoxon tests. 

```{r, lactobacillus_differ}
set.seed(42)
ps2 <-ps1
ps2 <- rarefy_even_depth(ps2,sample.size=14240, replace=FALSE, rngseed = 1) 
ps2 = subset_samples(ps2, cycle == "cycle_one"| cycle == "cycle_two"| cycle == "cycle_three_before_stress")
ps3 <- prune_taxa(taxa_sums(ps2) > 0, ps2)
abundance <- tax_glom(ps3, taxrank = 'Genus')
Lac = subset_taxa(abundance, Genus=="Lactobacillus")

#cycle 1
Chl <- subset_samples(Lac, treatment_cycle =="Control_cycle_1" | treatment_cycle =="Chlorothalonil_cycle_1")
Chl <- prune_taxa(taxa_sums(Chl) > 0, Chl)
melt<-psmelt(Chl)
wilcox.test(Abundance~treatment2, data=melt)
#W = 58, p-value = 0.4428

#cycle 2
Chl2 <- subset_samples(Lac, treatment_cycle =="Control_cycle_2" | treatment_cycle =="Chlorothalonil_cycle_2")
Chl2 <- prune_taxa(taxa_sums(Chl2) > 0, Chl2)
melt2<-psmelt(Chl2)
wilcox.test(Abundance~treatment2, data=melt2)
#W = 83, p-value = 0.5512

#cycle 3
Tet <- subset_samples(Lac, treatment_cycle=="Chlorothalonil_cycle_3_before_stress"|treatment_cycle=="control_Chlorothalonil_cycle_3_before_stress")
Tet <- prune_taxa(taxa_sums(Tet) > 0, Tet)
melt3<-psmelt(Tet)
wilcox.test(Abundance~treatment2, data=melt3)
#W = 42, p-value = 0.9314
```

```{r, Bartonella_differ}
Bart = subset_taxa(abundance, Genus=="Bartonella")

#cycle 1
Chl <- subset_samples(Bart, treatment_cycle =="Control_cycle_1" | treatment_cycle =="Chlorothalonil_cycle_1")
Chl <- prune_taxa(taxa_sums(Chl) > 0, Chl)
melt<-psmelt(Chl)
wilcox.test(Abundance~treatment2, data=melt)
#W = 83, p-value = 0.5512

#cycle 2
Chl2 <- subset_samples(Bart, treatment_cycle =="Control_cycle_2" | treatment_cycle =="Chlorothalonil_cycle_2")
Chl2 <- prune_taxa(taxa_sums(Chl2) > 0, Chl2)
melt2<-psmelt(Chl2)
wilcox.test(Abundance~treatment2, data=melt2)
#W = 58, p-value = 0.4428

#cycle 3
Tet <- subset_samples(Bart, treatment_cycle=="Chlorothalonil_cycle_3_before_stress"|treatment_cycle=="control_Chlorothalonil_cycle_3_before_stress")
Tet <- prune_taxa(taxa_sums(Tet) > 0, Tet)
melt3<-psmelt(Tet)
wilcox.test(Abundance~treatment2, data=melt3)
#W = 34, p-value = 0.6048
```
#### Gilliamella
```{r,Gilliamella_abund}
Gil = subset_taxa(abundance, Genus=="Gilliamella")

Chl <- subset_samples(Gil, treatment_cycle =="Control_cycle_1" | treatment_cycle =="Chlorothalonil_cycle_1")
Chl <- prune_taxa(taxa_sums(Chl) > 0, Chl)
melt<-psmelt(Chl)
wilcox.test(Abundance~treatment2, data=melt)
#W = 107, p-value = 0.0449

#cycle 2
Chl2 <- subset_samples(Gil, treatment_cycle =="Control_cycle_2" | treatment_cycle =="Chlorothalonil_cycle_2")
Chl2 <- prune_taxa(taxa_sums(Chl2) > 0, Chl2)
melt2<-psmelt(Chl2)
wilcox.test(Abundance~treatment2, data=melt2)
#W = 79, p-value = 0.7074

#cycle 3
Tet <- subset_samples(Gil, treatment_cycle=="Chlorothalonil_cycle_3_before_stress"|treatment_cycle=="control_Chlorothalonil_cycle_3_before_stress")
Tet <- prune_taxa(taxa_sums(Tet) > 0, Tet)
melt3<-psmelt(Tet)
wilcox.test(Abundance~treatment2, data=melt3)
#W = 33, p-value = 0.5457

```
#### Snodgrassella
```{r,Snodgrassella_abund}
Snod = subset_taxa(abundance, Genus=="Snodgrassella")

Chl <- subset_samples(Snod, treatment_cycle =="Control_cycle_1" | treatment_cycle =="Chlorothalonil_cycle_1")
Chl <- prune_taxa(taxa_sums(Chl) > 0, Chl)
melt<-psmelt(Chl)
wilcox.test(Abundance~treatment2, data=melt)
#W = 66, p-value = 0.7553

#cycle 2
Chl2 <- subset_samples(Snod, treatment_cycle =="Control_cycle_2" | treatment_cycle =="Chlorothalonil_cycle_2")
Chl2 <- prune_taxa(taxa_sums(Chl2) > 0, Chl2)
melt2<-psmelt(Chl2)
wilcox.test(Abundance~treatment2, data=melt2)
#W = 98.5, p-value = 0.1332

#cycle 3
Tet <- subset_samples(Snod, treatment_cycle=="Chlorothalonil_cycle_3_before_stress"|treatment_cycle=="control_Chlorothalonil_cycle_3_before_stress")
Tet <- prune_taxa(taxa_sums(Tet) > 0, Tet)
melt3<-psmelt(Tet)
wilcox.test(Abundance~treatment2, data=melt3)
#W = 48, p-value = 0.5457
```


#### Commensalibacter
```{r,Commensalibacter_abund}
Com = subset_taxa(abundance, Genus=="Commensalibacter")

Chl <- subset_samples(Com, treatment_cycle =="Control_cycle_1" | treatment_cycle =="Chlorothalonil_cycle_1")
Chl <- prune_taxa(taxa_sums(Chl) > 0, Chl)
melt<-psmelt(Chl)
wilcox.test(Abundance~treatment2, data=melt)
#W = 79, p-value = 0.7125

#cycle 2
Chl2 <- subset_samples(Com, treatment_cycle =="Control_cycle_2" | treatment_cycle =="Chlorothalonil_cycle_2")
Chl2 <- prune_taxa(taxa_sums(Chl2) > 0, Chl2)
melt2<-psmelt(Chl2)
wilcox.test(Abundance~treatment2, data=melt2)
#W = 98, p-value = 0.1432

#cycle 3
Tet <- subset_samples(Com, treatment_cycle=="Chlorothalonil_cycle_3_before_stress"|treatment_cycle=="control_Chlorothalonil_cycle_3_before_stress")
Tet <- prune_taxa(taxa_sums(Tet) > 0, Tet)
melt3<-psmelt(Tet)
wilcox.test(Abundance~treatment2, data=melt3)
#W = 40, p-value = 1
```

#### Bifidobacterium
```{r,Bifidobacterium_abund}
Bif = subset_taxa(abundance, Genus=="Bifidobacterium")

Chl <- subset_samples(Bif, treatment_cycle =="Control_cycle_1" | treatment_cycle =="Chlorothalonil_cycle_1")
Chl <- prune_taxa(taxa_sums(Chl) > 0, Chl)
melt<-psmelt(Chl)
wilcox.test(Abundance~treatment2, data=melt)
#W = 67, p-value = 0.7987

#cycle 2
Chl2 <- subset_samples(Bif, treatment_cycle =="Control_cycle_2" | treatment_cycle =="Chlorothalonil_cycle_2")
Chl2 <- prune_taxa(taxa_sums(Chl2) > 0, Chl2)
melt2<-psmelt(Chl2)
wilcox.test(Abundance~treatment2, data=melt2)
#W = 61, p-value = 0.5512

#cycle 3
Tet <- subset_samples(Bif, treatment_cycle=="Chlorothalonil_cycle_3_before_stress"|treatment_cycle=="control_Chlorothalonil_cycle_3_before_stress")
Tet <- prune_taxa(taxa_sums(Tet) > 0, Tet)
melt3<-psmelt(Tet)
wilcox.test(Abundance~treatment2, data=melt3)
#W = 28, p-value = 0.2973

```


### ASV

### have a look on most abundant ASVs in the core taxa to test for response variation on strain level

```{r Gilliamella_ASV}
tax<-ps1
tax = rarefy_even_depth(tax, rngseed=1, sample.size=14240, replace=F)
ps2 <- subset_samples(tax, cycle =="cycle_one"|cycle =="cycle_two"|cycle =="cycle_three_before_stress")
ps2 <- subset_samples(ps2, treatment2 =="Control"|treatment2 =="Chlorothalonil")
ps2 <- subset_taxa(ps2, Genus =="Gilliamella")

ps2.3 <- prune_taxa(taxa_sums(ps2) > 1000, ps2)
#ps2.4 <- subset_taxa(ps2.3, ASV !="ASV0068")

ps2m <- psmelt(ps2.3)

neworder <- c("cycle_one","cycle_two","cycle_three_before_stress")
change <- c("Cycle one","Cycle two","Cycle three")
ps_df_sum2 <- arrange(transform(ps2m, cycle=factor(cycle,levels=neworder,labels=change)),cycle)

p<-ggplot(data = ps_df_sum2, aes(x = treatment2, y = Abundance, colour = treatment2)) +
  geom_point(position = position_jitter())+geom_boxplot(show.legend=FALSE, alpha = 0,colour="grey53",size=0.7)+scale_color_manual(values=c("cyan4","#55596a"),guide = guide_legend(override.aes = list(shape = c(rep(15, 2)),size=7,alpha=0.5)))+theme_bw()
p<-p + facet_wrap(~ASV, scales="free_y",nrow=2)
p<-p + theme(strip.text = element_text(size=13,face="bold",color="grey40"))
taxa2<-p+theme(axis.title.y = element_text(size=13, face="bold"))+theme(axis.text.y = element_text(size=12, face="bold"))+theme(axis.text.x = element_blank())+theme(axis.title.x = element_blank())+theme(strip.background =element_rect(fill="white", color="white"))
taxa2<- taxa2 + theme(legend.title = element_blank(),legend.background = element_blank(),legend.key = element_blank(),legend.text = element_text(size=14, face="italic"))
taxa2+ggtitle("Gilliamella apicola")+theme(plot.title = element_text(hjust = 0.5, size=15, face="bold.italic",colour = "darkolivegreen"))+theme(axis.ticks.x=element_blank())
#ggsave("microbiome/figures/ASV_Gilliamella.png", height = 4.5, width = 8.5)
```


```{r Bifidobacterium_ASV}
tax<-ps1
tax = rarefy_even_depth(tax, rngseed=1, sample.size=14240, replace=F)
ps2 <- subset_samples(tax, cycle =="cycle_one"|cycle =="cycle_two"|cycle =="cycle_three_before_stress")
ps2 <- subset_samples(ps2, treatment2 =="Control"|treatment2 =="Chlorothalonil")
ps2 <- subset_taxa(ps2, Genus =="Bifidobacterium")

ps2.3 <- prune_taxa(taxa_sums(ps2) > 1000, ps2)
ps2m <- psmelt(ps2.3)

neworder <- c("cycle_one","cycle_two","cycle_three_before_stress")
change <- c("Cycle one","Cycle two","Cycle three")
ps_df_sum2 <- arrange(transform(ps2m, cycle=factor(cycle,levels=neworder,labels=change)),cycle)

p<-ggplot(data = ps_df_sum2, aes(x = treatment2, y = Abundance, colour = treatment2)) +
  geom_point(position = position_jitter())+geom_boxplot(show.legend=FALSE, alpha = 0,colour="grey53",size=0.7)+scale_color_manual(values=c("cyan4","#55596a"),guide = guide_legend(override.aes = list(shape = c(rep(15, 2)),size=7,alpha=0.5)))+theme_bw()
p<-p + facet_wrap(~ASV, scales="free_y",nrow=1)
p<-p + theme(strip.text = element_text(size=13,face="bold",color="grey40"))
taxa2<-p+theme(axis.title.y = element_text(size=13, face="bold"))+theme(axis.text.y = element_text(size=12, face="bold"))+theme(axis.text.x = element_blank())+theme(axis.title.x = element_blank())+theme(strip.background =element_rect(fill="white", color="white"))
taxa2<- taxa2 + theme(legend.title = element_blank(),legend.background = element_blank(),legend.key = element_blank(),legend.text = element_text(size=14, face="italic"))
taxa2+ggtitle("Bifidobacterium spp.")+theme(plot.title = element_text(hjust = 0.5, size=15, face="bold.italic",colour = "darkolivegreen"))+theme(axis.ticks.x=element_blank())
#ggsave("microbiome/figures/ASV_Bifidobacterium.png", height = 2.7, width = 8.5)
```

### species Lactobacillus
```{r Lacto_species}
tax<-ps1
tax = rarefy_even_depth(tax, rngseed=1, sample.size=14240, replace=F)
ps2 <- subset_samples(tax, cycle =="cycle_one"|cycle =="cycle_two"|cycle =="cycle_three_before_stress")
ps2 <- subset_samples(ps2, treatment2 =="Control"|treatment2 =="Chlorothalonil")
ps2 <- subset_taxa(ps2, Genus =="Lactobacillus")
ps2 <- prune_taxa(taxa_sums(ps2) > 1000, ps2)
ps2 <- tax_glom(ps2, taxrank = 'Species')
ps2m <- psmelt(ps2)

neworder <- c("cycle_one","cycle_two","cycle_three_before_stress")
change <- c("Cycle one","Cycle two","Cycle three")
ps_df_sum2 <- arrange(transform(ps2m, cycle=factor(cycle,levels=neworder,labels=change)),cycle)

p<-ggplot(data = ps_df_sum2, aes(x = treatment2, y = Abundance, colour = treatment2)) +
  geom_point(position = position_jitter())+geom_boxplot(show.legend=FALSE, alpha = 0,colour="grey53",size=0.7)+scale_color_manual(values=c("cyan4","#55596a"),guide = guide_legend(override.aes = list(shape = c(rep(15, 2)),size=7,alpha=0.5)))
p<-p + facet_wrap(~Species, scales="free_y",nrow=2)+theme_bw()
p<-p + theme(strip.text = element_text(size=13,face="bold.italic",color="grey40"))
taxa2<-p+theme(axis.title.y = element_text(size=13, face="bold"))+theme(axis.text.y = element_text(size=12, face="bold"))+theme(axis.text.x = element_blank())+theme(axis.title.x = element_blank())+theme(strip.background =element_rect(fill="white", color="white"))
taxa2<- taxa2 + theme(legend.title = element_blank(),legend.background = element_blank(),legend.key = element_blank(),legend.text = element_text(size=14, face="italic"))
taxa2+ggtitle("Lactobacillus spp.")+theme(plot.title = element_text(hjust = 0.5, size=15, face="bold.italic",colour = "darkolivegreen"))+theme(axis.ticks.x=element_blank())
#ggsave("microbiome/figures/Lactobacillus spp.png", height = 4.50, width = 8.5)
```


### ASV Lactobacillus
```{r ASV_L.apis}
tax<-ps1
tax = rarefy_even_depth(tax, rngseed=1, sample.size=14240, replace=F)
ps2 <- subset_samples(tax, cycle =="cycle_one"|cycle =="cycle_two"|cycle =="cycle_three_before_stress")
ps2 <- subset_samples(ps2, treatment2 =="Control"|treatment2 =="Chlorothalonil")
ps2 <- subset_taxa(ps2, Genus =="Lactobacillus")
ps2 <- subset_taxa(ps2, Species =="Lactobacillus apis")
ps2.3 <- prune_taxa(taxa_sums(ps2) > 1000, ps2)
ps2.3 <- prune_taxa(taxa_sums(ps2.3) > 0, ps2.3)
ps2m <- psmelt(ps2.3)
neworder <- c("cycle_one","cycle_two","cycle_three_before_stress")
change <- c("Cycle one","Cycle two","Cycle three")
ps_df_sum2 <- arrange(transform(ps2m, cycle=factor(cycle,levels=neworder,labels=change)),cycle)

p<-ggplot(data = ps_df_sum2, aes(x = treatment2, y = Abundance, colour = treatment2)) +
  geom_point(position = position_jitter())+geom_boxplot(show.legend=FALSE, alpha = 0,colour="grey53",size=0.7)+scale_color_manual(values=c("cyan4","#55596a"),guide = guide_legend(override.aes = list(shape = c(rep(15, 2)),size=7,alpha=0.5)))
p<-p + facet_wrap(~ASV, scales="free_y",nrow=1)+theme_bw()
p<-p + theme(strip.text = element_text(size=13,face="bold",color="grey40"))
taxa2<-p+theme(axis.title.y = element_text(size=13, face="bold"))+theme(axis.text.y = element_text(size=12, face="bold"))+theme(axis.text.x = element_blank())+theme(axis.title.x = element_blank())+theme(strip.background =element_rect(fill="white", color="white"))
taxa2<- taxa2 + theme(legend.title = element_blank(),legend.background = element_blank(),legend.key = element_blank(),legend.text = element_text(size=14, face="italic"))
taxa2+ggtitle("Lactobacillus apis")+theme(plot.title = element_text(hjust = 0.5, size=15, face="bold.italic",colour = "darkolivegreen"))+theme(axis.ticks.x=element_blank())
#ggsave("microbiome/figures/ASV_Lactobacillus apis2.png", height = 2.5, width = 6.1)
```
# Lactobacillus kullabergensis
```{r ASV_L.kullabergensis}
tax<-ps1
tax = rarefy_even_depth(tax, rngseed=1, sample.size=14240, replace=F)
ps2 <- subset_samples(tax, cycle =="cycle_one"|cycle =="cycle_two"|cycle =="cycle_three_before_stress")
ps2 <- subset_samples(ps2, treatment2 =="Control"|treatment2 =="Chlorothalonil")
ps2 <- subset_taxa(ps2, Genus =="Lactobacillus")
ps2 <- subset_taxa(ps2, Species =="Lactobacillus kullabergensis")
ps2.3 <- prune_taxa(taxa_sums(ps2) > 1000, ps2)
ps2.3 <- prune_taxa(taxa_sums(ps2.3) > 0, ps2.3)
ps2m <- psmelt(ps2.3)
neworder <- c("cycle_one","cycle_two","cycle_three_before_stress")
change <- c("Cycle one","Cycle two","Cycle three")
ps_df_sum2 <- arrange(transform(ps2m, cycle=factor(cycle,levels=neworder,labels=change)),cycle)

p<-ggplot(data = ps_df_sum2, aes(x = treatment2, y = Abundance, colour = treatment2)) +
  geom_point(position = position_jitter())+geom_boxplot(show.legend=FALSE, alpha = 0,colour="grey53",size=0.7)+scale_color_manual(values=c("cyan4","#55596a"),guide = guide_legend(override.aes = list(shape = c(rep(15, 2)),size=7,alpha=0.5)))
p<-p + facet_wrap(~ASV, scales="free_y",nrow=1)+theme_bw()
p<-p + theme(strip.text = element_text(size=13,face="bold",color="grey40"))
taxa2<-p+theme(axis.title.y = element_text(size=13, face="bold"))+theme(axis.text.y = element_text(size=12, face="bold"))+theme(axis.text.x = element_blank())+theme(axis.title.x = element_blank())+theme(strip.background =element_rect(fill="white", color="white"))
taxa2<- taxa2 + theme(legend.title = element_blank(),legend.background = element_blank(),legend.key = element_blank(),legend.text = element_text(size=15, face="italic"))
taxa2+ggtitle("L. kullabergensis")+theme(plot.title = element_text(hjust = 0.5, size=15, face="bold.italic",colour = "darkolivegreen"))+theme(axis.ticks.x=element_blank())
#ggsave("microbiome/figures/ASV_Lactobacillus kullabergensis2.png", height = 2.5, width = 6.5)
```
